Figures

### 2016-08-15 manuscript-main.Rmd
pushing edits so I can edit manuscript locally

---
title: "Aligning marine species range data to better serve science and conservation"
author: Casey O'Hara^1^, Jamie C. Afflerbach^1^, Courtney Scarborough^1^, Benjamin
  S. Halpern^1,2,3^
output:
  word_document: default
  pdf_document:
    number_sections: no
    toc: no
  html_document:
    highlight: haddock
    number_sections: no
    theme: spacelab
    toc: no
---

@@

``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}

@@

``` {r load data, echo = FALSE, warning = FALSE, message = FALSE}
### use data_explore.Rmd to create this file, if it does not yet exist.  This
### filters out presence = Extinct from IUCN cells.

@@

``` {r calc species numbers, echo = FALSE}
n_spp_am     <- nrow(spp_am)
n_spp_am_rev <- nrow(spp_am %>% filter(reviewed == 1))
n_spp_iucn   <- length(unique(spp_iucn_maps$iucn_sid))
n_spp_both   <- length(unique(spp_map_pairs$iucn_sid))
n_spp_all    <- n_spp_am + n_spp_iucn - n_spp_both

@@

``` {r abstract, child = 'ms_1_abst_sig.Rmd'}
```

@@

\\body

``` {r intro_results_discussion, child = 'ms_2_intro_res_disc.Rmd'}
```

@@

``` {r methods, child = 'ms_3_methods.Rmd'}
```

>>>>

---
title: "Aligning marine species range data to better serve science and conservation"
author: Casey O'Hara^1^, Jamie C. Afflerbach^1^, Courtney Scarborough^1^, Benjamin
  S. Halpern^1,2,3^
csl: refs/journal-of-applied-ecology.csl
output:
  word_document: default
  pdf_document:
    number_sections: no
    toc: no
  html_document:
    highlight: haddock
    number_sections: no
    theme: spacelab
    toc: no
bibliography: refs/zotero_lib.bib
---

@@

``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

@@

``` {r load data}
### use data_explore.Rmd to create this file, if it does not yet exist.  This
### filters out presence = Extinct from IUCN cells.

@@

``` {r calc species numbers}
n_spp_am     <- nrow(spp_am)
n_spp_am_rev <- nrow(spp_am %>% filter(reviewed == 1))
n_spp_iucn   <- length(unique(spp_iucn_maps$iucn_sid))
n_spp_both   <- length(unique(spp_map_pairs$iucn_sid))
n_spp_all    <- n_spp_am + n_spp_iucn - n_spp_both

@@

``` {r abstract, child = 'ms_0_abst_j_appl_ecol.Rmd'}
```

@@

``` {r intro, child = 'ms_1_intro.Rmd'}
```

``` {r methods, child = 'ms_2_methods.Rmd'}
```

``` {r results_discussion, child = 'ms_3_res_disc.Rmd'}
```

@@

### 2016-08-15 ms_1_intro.Rmd
pushing edits so I can edit manuscript locally

<!--
The two datasets ostensibly describe the same information, but significant differences in methodology and intent could lead to dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations. 
Discuss "truth" and uncertainty in species distributions.
There is an objective "truth"; 
-->
Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each presents one understanding of the "truth" - one database relying primarily on expert opinion and the other on model predictions.  Uncertainties inherent in the method and intent inevitably drive differences in the range predictions made by each dataset; significant differences could ___lead to, communicate, imply___ dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations.

Importantly, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.  Inaccurate indications of presence or absence could lead to ineffective marine reserve systems and management plans [@rondinini_tradeoffs_2006; @jetz_ecological_2008].

@@

It seems fair to assume that range predictions from each dataset would be more closely aligned for well-understood species.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align, determine several issues that lead to misalignment between predicted species distributions, and suggest methods to improve confidence in species range predictions.

@@

___we are not recommending one data set as better than the other!___

>>>>

___sentence about SAUP - how can we rephrase the "two global-scale repositories" sentence to exclude it? "broad taxonomic coverage"?  something about not focusing just on economically important species? (is that a valid assumption about SAUP)___

___or: "Several global-scale repositories predict ... In this paper we focus on two datasets that cover a broad taxonomic range: ..."___

<!--
The two datasets ostensibly describe the same information, but significant differences in methodology and intent could lead to dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations. 
Discuss "truth" and uncertainty in species distributions.
There is an objective "truth"; 
-->
Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each presents one understanding of the truth - one database relying primarily on expert opinion and the other on modeling.  Uncertainties inherent in the methods and intents inevitably drive differences in the range predictions made by each dataset.  Significant differences could ___lead to, communicate, imply___ dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations.  Importantly, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.  Inaccurate indications of presence or absence could lead to ineffective marine reserve systems and management plans [@rondinini_tradeoffs_2006; @jetz_ecological_2008].

@@

We can reasonably assume that range predictions from these two datasets would be more tightly aligned for well-understood species.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the predicted ranges align in distribution and area, and determine several issues that lead to poor alignment between the two datasets. ___

@@

___While we cannot recommend one data set as more accurate than the other, we do suggest methods to improve confidence in species range predictions and conservation outcomes based on these two datasets.___

### 2016-08-15 ms_2_methods.Rmd
pushing edits so I can edit manuscript locally

The IUCN publishes species range maps based on expert input of spatial boundaries of a given species' "limits of distribution" [@iucn_iucn_2015] - essentially a refined extent of occurrence (EOO), based on observation records and informed by expert understanding of species' range and habitat preferences.  In contrast, AquaMaps models species distribution based on environmental preferences (e.g., temperature, depth, salinity) deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a 0.5 degree grid, creating a global map of "probability of occurrence" for each species [@kaschner_mapping_2006; @ready_predicting_2010].  Studies using AquaMaps data frequently define "presence" by applying a probability threshold, e.g., the Ocean Health Index defines presence as 40% or greater probability of occurrence [@halpern_index_2012; @halpern_patterns_2015].

@@

__Comparison of taxonomic and regional distribution__:
To examine the overall taxonomic distribution across the spatial datasets (Fig. 1A), we grouped species by taxonomic class and data source (IUCN, AquaMaps, or both), and examined the proportion of each class represented in each data source category.

To compare the spatial representation of the two datasets directly, we rasterized the IUCN species polygons to the same 0.5° grid as the AquaMaps species maps; we determined species presence within a grid cell as any non-zero overlap of a species polygon with the cell.  For the AquaMaps dataset, we determined per-cell species count by including all species with non-zero probability of occurrence, to best approximate the "extent of occurrence" generally indicated by IUCN maps.  We represented relative distribution of species count for each dataset by plotting average species count against latitude and longitude (Fig. 1B, 1C).

__Comparison of paired maps__:
Using genus and species binomials as a matching key, we selected the subset of marine species that have range maps in both IUCN and AquaMaps current native distribution.  ___To avoid double-counting, we removed subpopulations and species aliases. Used the taxize package? which version?___  We determined species presence within each spatial cell using the same criteria outlined above.

@@

For each species with ranges described in both IUCN and AquaMaps, $A_{small}$ and $A_{large}$ indicate the smaller and larger range representation (regardless of which dataset).  $A_{small \cap large}$ represents the amount of overlapping area between the two datasets.

@@

Dividing the plot into quadrants based on mean values along each axis, we roughly categorized four qualities of alignment, and for each taxonomic group, based on IUCN taxonomic groupings, determined the proportion of species that fell into each quality category (2B).

Noting that a large proportion of corals were identified in the upper-left quadrant, we examined a handful of IUCN coral maps and found that each predicted coral presence in far deeper waters than would be expected based on the IUCN's habitat description.  Using the IUCN's API [REF] to find life history information for all IUCN-mapped corals, we found that none of the paired-map corals is described as occurring deeper than 200 m, ___and most are found in far shallower waters.___  We created a 200 m bathymetry raster from Natural Earth's 200 m bathymetry polygon [@natural_earth_XXX] and masked our IUCN coral rasters to exclude presence below 200 m.  The resulting maps were again compared to the AquaMaps rasters; ___figure XXXX___ plots shifts in coral species alignment due to this depth clipping. ___does some of this belong in the results/discussion?___

**Methods for Ocean Health Index (OHI) case study:**
The Species component of the OHI Biodiversity goal (SPP) uses IUCN Red List extinction risk as a proxy for a species' health.  Each species is assigned a risk score based on IUCN Red List extinction risk category, from Least Concern (0.0) to Extinct (1.0) (Data Deficient and unassessed species are excluded).  Species range is determined from its IUCN range map, where available, or its AquaMaps range map ("presence" defined as a 40% or greater probability of occurrence) if an IUCN map is unavailable.  For each assessed OHI region, an area-weighted mean extinction risk across all species is calculated, and rescaled to calculate a status score from 0 (mean extinction risk = 0.75) to 100 (mean extinction risk = 0) [@halpern_index_2012; @halpern_patterns_2015].

>>>>

The IUCN publishes species range maps based on expert input of spatial boundaries of a given species' "limits of distribution" [@iucn_iucn_2015] - essentially a refined extent of occurrence (EOO), based on observation records and informed by expert understanding of species' range and habitat preferences.  In contrast, AquaMaps models species distribution based on environmental preferences (e.g., temperature, depth, salinity) deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a 0.5 degree grid, creating a global map of "probability of occurrence" for each species [@kaschner_mapping_2006; @ready_predicting_2010].  Studies using AquaMaps data frequently define "presence" by applying a probability threshold, e.g., previously, the Ocean Health Index defined presence as 40% or greater probability of occurrence [@halpern_index_2012; @halpern_patterns_2015].  In the 2016 OHI assessment, the presence threshold was eliminated, to better approximate the "extent of occurrence" generally indicated by IUCN maps, since the two datasets are used together. [@REF???]

@@

__Comparison of taxonomic and regional distribution__:
To examine the overall taxonomic distribution across the spatial datasets (Fig. 1A), we grouped species by taxonomic class and data source, and determined the proportion of each class represented in each dataset.  We defined "paired map" species as those with spatial data in both datasets.

To compare the spatial representation of the two datasets directly, we rasterized the IUCN species polygons to the same 0.5° grid as the AquaMaps species maps; we determined species presence within a grid cell as any non-zero overlap of a species polygon with the cell  ___(Fig. SXXX)___.  For the AquaMaps dataset, we determined per-cell species count by including all species with non-zero probability of occurrence, to best approximate the "extent of occurrence" generally indicated by IUCN maps ___(Fig. SXXX)___.  We represented relative distribution of species count for each dataset by plotting average species count against latitude and longitude (Fig. 1B, 1C).

__Comparison of paired maps__:
Using genus and species binomials as a matching key, we selected the "paired map" species - the subset of marine species that have range maps in both IUCN and AquaMaps current native distribution.  _We used the taxize package [@REF] in R [@REF] to standardize species names and synonyms; for species with separate subpopulation maps in IUCN, we combined all subpopulations to create a single global population._  ___To avoid double-counting, we removed subpopulations and species aliases. Used the taxize package? which version?___  We determined species presence within each spatial cell using the same criteria outlined above.

@@

For each paired map species, $A_{small}$ and $A_{large}$ indicate the smaller and larger range representation (regardless of which dataset).  $A_{small \cap large}$ represents the amount of overlapping area between the two datasets.

@@

Dividing the plot into quadrants based on mean values along each axis, we roughly categorized four qualities of alignment, and for each taxonomic group, based on IUCN taxonomic groupings, determined the proportion of paired map species that fell into each quality category (2B).

Noting that a large proportion of corals were identified in the upper-left quadrant, we examined a handful of IUCN coral maps and found that each predicted coral presence in far deeper waters than would be expected based on the IUCN's habitat description.  Using the IUCN's API [REF] to find life history information for all IUCN-mapped corals, we found that none of the paired-map corals is described as occurring deeper than 200 m; 94% are noted as being limited to 50 m or shallower.  We created a 200 m bathymetry raster from Natural Earth's 200 m bathymetry polygon [@natural_earth_XXX] ___(Fig. SXXX)___ and masked our IUCN coral rasters to exclude presence below 200 m.  The resulting maps were again compared to the AquaMaps rasters; ___Fig. XXXX___ plots shifts in coral species alignment due to this depth clipping. ___does some of this belong in the results/discussion?___

___To identify data-poor species, we used the robis??? and rgbif??? packages in R to identify known occurrences of each species; occurrences were averaged between the two bases as a proxy for data richness.___

**Methods for Ocean Health Index (OHI) case study:**
The Species component of the OHI Biodiversity goal (SPP) uses IUCN Red List extinction risk as a proxy for a species' health.  Each species is assigned a risk score based on IUCN Red List extinction risk category, from Least Concern (0.0) to Extinct (1.0) (Data Deficient and unassessed species are excluded).  Species range is determined from its IUCN range map, where available, or its AquaMaps range map if an IUCN map is unavailable.  For each assessed OHI region, an area-weighted mean extinction risk across all species is calculated, and rescaled to calculate a status score from 0 (mean extinction risk = 0.75) to 100 (mean extinction risk = 0) [@halpern_index_2012; @halpern_patterns_2015].

_To highlight the differences in the datasets, we recalculated the OHI SPP status values using only one complete dataset at a time.  For the AquaMaps dataset, we used a presence threshold of 0% to align with the OHI 2016 assessment._  ___Fig. 3 compares these single-dataset scores to the 2016 OHI SPP status values.___

### 2016-08-15 ms_5_figures.Rmd
pushing edits so I can edit manuscript locally

![](figures/fig1abc_am_iucn_maps.png)

@@

![](figures/fig2ab_quadmap_sppbars.png)

@@

![Absolute change in SPP score](figures/fig3_boxplot_abs_diff.png)

@@

>>>>

![](figures/fig1a_bar_by_taxa_all.png)
![](figures/fig1b_all_aquamaps_map.png)
![](figures/fig1c_all_iucn_map.png)

@@

![](figures/fig2a_scatter_quadmap.png)
![](figures/fig2b_barchart_spp_gp_quads.png)

@@

<a id = 'fig3'></a>

![paired map comparison](figures/fig3a_violinplot_paired_maps.png)

why is the global one not showing up?:
![global full species list comparison](figures/fig3a_violinplot_ohi_global.png)

@@

### New figures - corals

![coral area change due to depth clipping](figures/fig5a_coral_area_plot.png)

![coral quadmap change due to depth clipping](figures/fig5b_scatter_quadmap_corals.png)
